---
title: "Tools for Inspecting Metabolomic Batch Effects and Implementing QC-Based Batch Corrections"
author: "Erick Helmeczi"
date: "`r Sys.Date()`"
output: html_document
runtime: shiny
---
<hr style="border:1px solid black">

<div align = "center">

## Application Summary

</div>

<div align = "left">

<hr style="border:1px solid black">

This interactive R Markdown document provides tools and summaries to help students, researchers, and technicians identify batch effects in their metabolomic data sets, optimize appropriate batch correction algorithms, and export corrected matrices. These features are separated into the following three sections: 

  1. A data summary prior to the application of any batch correction. This includes a table summary highlighting key statistics for each metabolite and a principal component analysis (PCA) to help compare study technical and biological variations. Furthermore, the inclusion of control charts help identify intra- and inter-batch effects. The main purpose of this section is to help identify if batch corrections are necessary and to establish the baseline variation of the data which will hopefully be reducible in section 2.
  2. Several batch correction algorithms are provided along with comparative figures and statistics to help users quantify the effectiveness of the batch correction algorithms. This includes many of the same tools in section 1 so direct comparisons can be made between pre- and post-corrected data.
  3. Lastly, an export feature is available for the batch corrected data set. Normalization transformers and feature filters can also be applied to the data before export.

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

```{r, include = FALSE}

# Load packages
if (!require("pacman", quietly = TRUE)) install.packages("pacman")
if (!require("BiocManager", quietly = TRUE)) install.packages("BiocManager")

pacman::p_load("tidyverse", "stats", "rmarkdown", "factoextra", "pmp",
               install = TRUE)

```

```{r}
# Read and prepare data
raw_data <- read.csv("study matrix.csv")
raw_data$class <- as.factor(raw_data$class)
raw_data$batch <- as.factor(raw_data$batch)

```

```{r}

## Functions ===============================================================================================

# Transform Function (i.e. log-transform)

transform_function <- function(input_data, transform){
  
  temp_df <- input_data
  
  if (transform == "Log10") {
    temp_df[, 5:ncol(temp_df)] <- log(temp_df[, 5:ncol(temp_df)], base = 10)
  } else if (transform == "ln") {
    temp_df[, 5:ncol(temp_df)] <- log(temp_df[, 5:ncol(temp_df)])
  } else if (transform == "Cube Root") {
    temp_df[, 5:ncol(temp_df)] <- (temp_df[, 5:ncol(temp_df)])^(1/3)
  }
  
  return(temp_df)
  
}

# Auto-Scale Function

autoscale_function <- function(input_data, autoscale){
  
  temp_df <- input_data
  
  if(autoscale == TRUE){
    temp_df[, 5:ncol(temp_df)] <- scale(temp_df[, 5:ncol(temp_df)], center = TRUE, scale = TRUE)
  }else{
    temp_df
  }
  
  return(temp_df)
  
}

```

```{r}

## Reactive Functions ==================================================================================

# Correction Algorithm (QC-RSC)

qcrsc_corrected <- reactive({
  
  df <- raw_data
  
  results_temp <- QCRSC(df = df[, 5:ncol(df)],
                            order = df$run_order,
                            batch = df$batch,
                            classes = df$class,
                            spar = input$smoothing_parameter,
                            log = FALSE,
                            minQC = 4,
                            qc_label = "QC")
  
  results_temp <- t(results_temp)
  
  df[, 5:ncol(df)] <- results_temp
  
  return(df)
  
})

# Correction Algorithm (QC-MN)

qcmn_corrected <- reactive({
  
  df <- raw_data
  
  # perform batch-wise
  
  for(b in 1:length((unique(df$batch)))){
    
    df_temp <- subset(df, df$batch == unique(df$batch)[b])
    qc_df <- subset(df_temp, df_temp$class == "QC")
    
      for (i in 1:nrow(df_temp)){
    
        # find nearest QC indexes
        qc_indexes <- sort(abs(qc_df$run_order - df_temp$run_order[i]))[1:input$n_qc] + df_temp$run_order[i]
        qc_indexes <- which(qc_df$run_order %in% qc_indexes)
    
        # get metabolite medians
        qc_medians <- apply(qc_df[qc_indexes, 5:ncol(qc_df)], 2, median)
    
        # normalize data to medians
        df[which(df$pbm_id == df_temp$pbm_id[i]), 5:ncol(df)] <- df[which(df$pbm_id == df_temp$pbm_id[i]), 5:ncol(df)]  / qc_medians
    
      }
  }

  # adjust final scale
  df[, 5:ncol(df)] <- Map(`*`, df[, 5:ncol(df)], colMeans(raw_data[, 5:ncol(raw_data)]))
  
  return(df)
  
})


```

```{r}

## Plots =======================================================================================================

# Control Chart ------------------------------------------------------------------------------------------------

create_control_chart <- function(data, transform_input, metabolite_input, class_filter_input, point_size_input, corrected_data) {
  
  # Prepare data
  
  plot_data <- data
  
  ## get batch corrected data
  if (corrected_data == TRUE){
    plot_data <- if (input$batch_correction == "QC-RSC") {
    qcrsc_corrected()
    } else if (input$batch_correction == "QC-MN") {
    qcmn_corrected()
    }
  }
  
  # transform data
  plot_data <- transform_function(input_data = plot_data, transform = transform_input)
  
  # filter data
  plot_data <- if (class_filter_input != "Both") {
    subset(plot_data, plot_data$class == class_filter_input)
  } else {
    plot_data
  }
  
  # Calculate metrics for control chart aesthetics
  mean_value <- mean(plot_data[[metabolite_input]])
  upper_limit <- mean(plot_data[[metabolite_input]]) + 1.96 * sd(plot_data[[metabolite_input]])
  lower_limit <- mean(plot_data[[metabolite_input]]) - 1.96 * sd(plot_data[[metabolite_input]])
  
  # Create the plot
  ggplot(data = plot_data, aes(x = run_order, y = .data[[metabolite_input]], color = batch)) +
    geom_point(size = point_size_input) +
    geom_hline(yintercept = mean_value, linetype = "solid", color = "black", size = 0.5) +
    geom_hline(yintercept = upper_limit, linetype = "dashed", color = "red", size = 0.5) +
    geom_hline(yintercept = lower_limit, linetype = "dashed", color = "red", size = 0.5) +
    theme_classic() +
    theme(text = element_text(size = 25),
          axis.text = element_text(color="black"),
          axis.ticks = element_line(color = "black")) +
    scale_x_continuous(breaks = scales::breaks_pretty(), name = "Run Order") +
    scale_y_continuous(breaks = scales::breaks_pretty(n = 8), name = "Response") +
    labs(color = "Batch") +
    ggtitle(paste(metabolite_input, "Control Chart", sep = " "))
}

# PCA ----------------------------------------------------------------------------------------------------------

create_PCA <- function(data, autoscale_input, transform_input, coloring_scheme_input, point_size_input, opacity_input, sample_label_input, corrected_data) {
  
  plot_data <- data
  
  ## get batch corrected data
  if (corrected_data == TRUE){
    plot_data <- if (input$batch_correction == "QC-RSC") {
    qcrsc_corrected()
    } else if (input$batch_correction == "QC-MN") {
    qcmn_corrected()
    }
  }
  
  # transform data
  plot_data <- transform_function(input_data = plot_data, transform_input)
  # Scale data
  pca_data <- autoscale_function(input_data = plot_data, autoscale_input)
  
  # convert class and batch to numeric for prcomp function
  pca_data$class <- ifelse(pca_data$class == "Sample", 1, 0)
  pca_data$batch <- is.numeric(pca_data$batch)
 
  # determine coloring scheme
  index <- ifelse(coloring_scheme_input == "Class", 2, 3)
  
  # compute PCA
  pca_results <- prcomp(pca_data[, c(5:ncol(pca_data))], 
                 center = FALSE,
                 scale = FALSE)
 
  scree.plot <- fviz_eig(pca_results)

  pc1 <- scree.plot[["data"]][["eig"]][1] %>%
    round(., 1)
  pc2 <- scree.plot[["data"]][["eig"]][2] %>%
    round(., 1)
  
  # plot
  fviz_pca_ind(pca_results,
             col.ind = data[, index],
             addEllipses = TRUE, 
             legend.title = colnames(data)[index],
             label = "none",
             repel = FALSE,
             mean.point = FALSE,
             pointsize = point_size_input,
             pointshape = 19,
             alpha = opacity_input) +
  theme(text = element_text(size = 20, family = "sans"),
        plot.title = element_blank(),
        axis.text = element_text(colour="black", face = "plain", size = 20, family = "sans"),
        axis.line = element_line(linewidth = 0.5, colour = "black", linetype = 1)) +
  scale_y_continuous(name = paste("PC 2 (", pc2, "%)", sep = ""),
                     breaks = scales::pretty_breaks(n = 10)) +
  scale_x_continuous(name = paste("PC 1 (", pc1, "%)", sep = ""),
                     breaks = scales::pretty_breaks(n = 10)) +
  if (sample_label_input == TRUE){
      geom_text(aes(label = data$pbm_id), vjust = -1)
  }else{
      geom_text(aes(label = ""), vjust = -1)
  }

}

# Histogram ---------------------------------------------------------------------------------------------

create_histogram <- function (data, transform_input, metabolite_input, bin_input, class_filter_input, corrected_data){
  
  plot_data <- data
  
  # get batch corrected data
  if (corrected_data == TRUE){
    plot_data <- if (input$batch_correction == "QC-RSC") {
    qcrsc_corrected()
    } else if (input$batch_correction == "QC-MN") {
    qcmn_corrected()
    }
  }
  
  # transform data
  plot_data <- transform_function(input_data = plot_data, transform_input)
  
  # filter data
  plot_data <- if (class_filter_input != "Both") {
    subset(plot_data, plot_data$class == class_filter_input)
  } else {
    plot_data
  }

  # plot
  ggplot(data = plot_data, aes(x = .data[[metabolite_input]])) +
    geom_histogram(bins = bin_input, color = "white", alpha = 0.7, aes(fill = batch)) +
    theme_classic() +
    theme(text = element_text(size = 25),
          axis.text = element_text(color="black"),
          axis.ticks = element_line(color = "black")) +
    scale_x_continuous(breaks = scales::breaks_pretty(), name = paste(input$metabolite, "Response", sep = " ")) +
    scale_y_continuous(breaks = scales::breaks_pretty(), name = "Count") +
    ggtitle(paste(input$metabolite, "Histogram", sep = " "))
  
}

# Q-Q Plot ----------------------------------------------------------------------------------------------------------

create_qq <- function (data, transform_input, metabolite_input, class_filter_input, corrected_data){
  
  plot_data <- data
  
  # get batch corrected data
  if (corrected_data == TRUE){
    plot_data <- if (input$batch_correction == "QC-RSC") {
    qcrsc_corrected()
    } else if (input$batch_correction == "QC-MN") {
    qcmn_corrected()
    }
  }
  
  # transform data
  plot_data <- transform_function(input_data = plot_data, transform_input)
  
  # filter data
  plot_data <- if (class_filter_input != "Both") {
    subset(plot_data, plot_data$class == class_filter_input)
  } else {
    plot_data
  }

  # plot
  ggplot(data = plot_data, aes(sample = .data[[metabolite_input]])) +
    stat_qq() +
    stat_qq_line() +
    theme_classic() + 
    theme(text = element_text(size = 25),
          axis.text = element_text(color="black"),
          axis.ticks = element_line(color = "black")) +
    scale_x_continuous(breaks = scales::breaks_pretty(), name = "Normal Theoretical Quantiles") +
    scale_y_continuous(breaks = scales::breaks_pretty(), paste(metabolite_input, "Response", sep = " ")) +
    ggtitle(paste(metabolite_input, "Q-Q Plot", sep = " "))
  
}

# Volcano Plot --------------------------------------------------------------------------------------------

create_volcano <- function (data, transform_input, class_filter_input, point_size_input, opacity_input, sample_label_input){
  
  data_uncorrected <- data
  data_corrected <- data
  
  # apply correction
  data_corrected <- if (input$batch_correction == "QC-RSC") {
  qcrsc_corrected()
  } else if (input$batch_correction == "QC-MN") {
  qcmn_corrected()
  }
  
  # transform data
  data_corrected <- transform_function(input_data = data_corrected, transform_input)
  data_uncorrected <- transform_function(input_data = data_uncorrected, transform_input)
  
  # filter data
  data_corrected <- if (class_filter_input != "Both") {
    subset(data_corrected, data_corrected$class == class_filter_input)
  } else {
    data_corrected
  }
  
  data_uncorrected <- if (class_filter_input != "Both") {
  subset(data_uncorrected, data_uncorrected$class == class_filter_input)
  } else {
  data_uncorrected
  }
  
  # compute cvs
  corrected_cvs <- apply(data_corrected[, 5:ncol(data_corrected)], 2, sd) / colMeans(data_corrected[, 5:ncol(data_corrected)]) * 100
  uncorrected_cvs <- apply(data_uncorrected[, 5:ncol(data_uncorrected)], 2, sd) / colMeans(data_uncorrected[, 5:ncol(data_corrected)]) * 100
  
  # summarize plotting data
  plot_data <- data.frame("metabolite" = colnames(data[5:ncol(data)]),
                          "cv_fold_change" = uncorrected_cvs / corrected_cvs,
                          "cv" = corrected_cvs)
  plot_data$group <- ifelse(plot_data$cv_fold_change < 1, 0, 1) %>%
    as.factor()
  
  # Make a list to store outputs
  volcano_list <- list()

  # plot
  p <- ggplot(data = plot_data, aes(x = cv_fold_change, y = cv, color = group)) +
    geom_point(size = point_size_input, alpha = opacity_input) +
    theme_classic() + 
    scale_color_manual(values = c("#00BA38", "#F8766D")) +
    theme(text = element_text(size = 25),
          axis.text = element_text(color="black"),
          axis.ticks = element_line(color = "black"),
          legend.position = "none") +
    scale_x_continuous(breaks = scales::breaks_pretty(n = 10), name = "CV Fold-Change", expand = c(0.2, 0.2)) +
    scale_y_continuous(breaks = scales::breaks_pretty(n = 10), name = "Post-Correction CV (%)", expand = c(0.2, 0.2)) +
  if (sample_label_input == TRUE){
      geom_text(aes(label = plot_data$metabolite), vjust = -1, size = point_size_input)
  }else{
      geom_text(aes(label = ""), vjust = -1)
  }
  
  volcano_list[[1]] <- p
  
  volcano_list[[2]] <- paste("Following batch correction the median CV fold-change was ", round(median(plot_data$cv_fold_change), digits = 2),"-fold with an IQR of ", round(quantile(plot_data$cv_fold_change)[2], digits  = 2), " to ", round(quantile(plot_data$cv_fold_change)[4], digits = 2), "-fold. ", "The new median CV is ", round(median(plot_data$cv), digits = 2), "% with an IQR of ", round(quantile(plot_data$cv)[2], digits = 2), "-", round(quantile(plot_data$cv)[4], digits = 2), "%.", sep = "")
  
  return(volcano_list)
  
}

## Tables ==================================================================================================

create_table <- function (data, transform_input, corrected_data){
  
  table_data <- data
  
  # get batch corrected data
  if (corrected_data == TRUE){
    table_data <- if (input$batch_correction == "QC-RSC") {
    qcrsc_corrected()
    } else if (input$batch_correction == "QC-MN") {
    qcmn_corrected()
    }
  }
  
  # transform data
  table_data <- transform_function(input_data = table_data, transform_input)

  # subset data
  sample_df <- table_data[which(table_data$class == "Sample"), 5:ncol(table_data)]
  qc_df <- table_data[which(table_data$class == "QC"), 5:ncol(table_data)]
  
  # calculate CVs
  biological_cv <- apply(sample_df, 2, sd) / colMeans(sample_df) * 100
  technical_cv <- apply(qc_df, 2, sd) / colMeans(qc_df) * 100
  
  # generate table summary

  table_summary <- data.frame("Metabolite" = colnames(sample_df),
                            "Mean" = colMeans(sample_df),
                            "Standard_Deviation" = apply(sample_df, 2, sd),
                            "Biological_CV" = biological_cv,
                            "Technical_CV" = technical_cv,
                            "ICC" = (apply(sample_df, 2, sd))^2 / ((apply(sample_df, 2, sd))^2 + (apply(qc_df, 2, sd))^2))

  table_summary[,2:ncol(table_summary)] <- round(table_summary[,2:ncol(table_summary)], digits = 2)
 
  table_list <- list()
  
  table_list[[1]] <- table_summary

  statement <- paste("In summary, the median technical CV is ", round(median(table_summary$Technical_CV), digits = 2), "% with a range of ", round(range(table_summary$Technical_CV), digits = 2)[1], " - " , round(range(table_summary$Technical_CV), digits = 2)[2], "%. The median biological CV is ", round(median(table_summary$Biological_CV), digits = 2),  "% with a range of ", round(range(table_summary$Biological_CV), digits = 2)[1], " - " , round(range(table_summary$Biological_CV), digits = 2)[2], "%", sep = "")
  
  table_list <- append(table_list, statement)
  
  return(table_list)
   
}

### Filtering ============================================================================================

filter_function <- function(data, transform_input, corrected_data, qc_cv_threshold, icc_threshold){
  
  filtered_data <- data
  
  # get batch corrected data
  if (corrected_data == TRUE){
    filtered_data <- if (input$batch_correction == "QC-RSC") {
    qcrsc_corrected()
    } else if (input$batch_correction == "QC-MN") {
    qcmn_corrected()
    }
  }
  
  # transform data
  filtered_data <- transform_function(input_data = filtered_data, transform_input)

  # subset data
  sample_df <- filtered_data[which(filtered_data$class == "Sample"), 5:ncol(filtered_data)]
  qc_df <- filtered_data[which(filtered_data$class == "QC"), 5:ncol(filtered_data)]
  
  # calculate technical CV
  technical_cv <- apply(qc_df, 2, sd) / colMeans(qc_df) * 100
  
  # Calculate ICC
  icc <- (apply(sample_df, 2, sd))^2 / ((apply(sample_df, 2, sd))^2 + (apply(qc_df, 2, sd))^2)
  
  #metabolite outside cv threshold
  cv_metabolites <- colnames(qc_df)[which(technical_cv > qc_cv_threshold)]
  
  #metabolite outside icc threshold
  icc_metabolites <- colnames(qc_df)[which(icc < icc_threshold)]
  
  # save outputs to list
  filtering_list <- list()
  filtering_list[[1]] <- ifelse(length(cv_metabolites) == 0, "None", paste(cv_metabolites, collapse=', ' ))
  filtering_list[[2]] <- ifelse(length(icc_metabolites) == 0, "None", paste(icc_metabolites, collapse=', ' ))
  
  return(filtering_list)
}

### Exporting ============================================================================================

export_function <- function(data, transform_input, corrected_data, qc_cv_threshold, icc_threshold, csv_name){
  
  export_data <- data
  
  # get batch corrected data
  if (corrected_data == TRUE){
    export_data <- if (input$batch_correction == "QC-RSC") {
    qcrsc_corrected()
    } else if (input$batch_correction == "QC-MN") {
    qcmn_corrected()
    }
  }
  
  # transform data
  export_data <- transform_function(input_data = export_data, transform_input)

  # subset data
  sample_df <- export_data[which(export_data$class == "Sample"), 5:ncol(export_data)]
  qc_df <- export_data[which(export_data$class == "QC"), 5:ncol(export_data)]
  
  # calculate technical CV
  technical_cv <- apply(qc_df, 2, sd) / colMeans(qc_df) * 100
  
  # Calculate ICC
  icc <- (apply(sample_df, 2, sd))^2 / ((apply(sample_df, 2, sd))^2 + (apply(qc_df, 2, sd))^2)
  
  #metabolite outside cv threshold
  cv_metabolites <- colnames(qc_df)[which(technical_cv > qc_cv_threshold)]
  
  #metabolite outside icc threshold
  icc_metabolites <- colnames(qc_df)[which(icc < icc_threshold)]
  
  # metabolites to remove
  remove_met <- unique(c(cv_metabolites, icc_metabolites))
  
  #subset export data
  export_data <- export_data[, !(colnames(export_data) %in% remove_met)]
  
  file_name <- paste(csv_name, ".csv", sep = "")
  
  write.csv(export_data, file = file_name, row.names = FALSE)
}

```
  
<hr style="border:1px solid black">

<div align = "center">

## Section 1: Data Summary Before Batch Correction

</div>

<div align = "left">

<hr style="border:1px solid black">

---

### Data Transformation

Each section provides users with an opportunity to select a data transformation which is only applied to that section (i.e. the transformation selected here will only be applied to section 1). Section 3 evaluates on the impact of the transformation on data distribution so an appropriate transformation can be selected before data export.

```{r}

inputPanel(
  
  selectInput(
  inputId = "transformer_section_1",
  label = "Data Transformer:",
  choices = c("None", "Log10", "ln", "Cube Root"),
  selected = "None",
  multiple = FALSE,
  selectize = TRUE,
  width = NULL,
  size = NULL)
  
  
)

```

### Table Summary

The table below summarizes key statistics for each metabolite. The columns "Mean" and "Standard_Deviation" were calculated with sample data (QC data is omitted). Biological and technical precisions have units of percent. The intraclass correlation coefficient (ICC) was calculated with the following equation:

$$
\Large ICC = \frac{\sigma^2_{\text{Biological}}} {\sigma^2_{\text{Biological}} + \sigma^2_{\text{Technical}}}
$$


```{r}

# Input Table

renderDataTable({
  
create_table(data = raw_data, transform_input = input$transformer_section_1, corrected_data = FALSE)[[1]]
  
})


renderText({
  
  create_table(data = raw_data, transform_input = input$transformer_section_1, corrected_data = FALSE)[[2]][1]
  
})


```

---

### Principal Component Analysis

Principal component analysis (PCA) is often employed to help summarize technical and biological precision in metabolomic studies where the number of metabolites are too great to be summarized individually. In the absence of batch effects, QC samples should be clustered tightly together as their variation should appear small compared to the biological variation. However, in the presence of batch effects, the spread of the QC samples typically increases due to the higher technical variation. Setting "Color BY:" to "Class" will allow users to explore their data in this sense. However, "Color BY:" can also be set to "Batch" in which case points are colored by batch. Here, the presence of clusters with little overlap indicate batch effects are the main source of variation in the data. When performing PCA it is highly recommended to autoscale data as variable weights will be strongly impacted by their absolute values.

```{r}

inputPanel(
  
    selectInput(
    inputId = "autoscaler",
    label = "Autoscale Data:",
    choices = c("TRUE", "FALSE"),
    selected = "TRUE",
    multiple = FALSE,
    selectize = TRUE,
    width = NULL,
    size = NULL),
    
    selectInput(
    inputId = "coloring_scheme_1",
    label = "Color By:",
    choices = c("Class", "Batch"),
    selected = "Class",
    multiple = FALSE,
    selectize = TRUE,
    width = NULL,
    size = NULL),
  
    selectInput(
    inputId = "sample_label_pca_1",
    label = "Sample Labels:",
    choices = c("TRUE", "FALSE"),
    selected = "FALSE",
    multiple = FALSE,
    selectize = TRUE,
    width = NULL,
    size = NULL),
  
  sliderInput(
    inputId = "point_size_pca_1", 
    label = "Point Size:",
    min = 1, 
    max = 10,
    value = 3),
  
    sliderInput(
    inputId = "opacity_pca_1", 
    label = "Opacity:",
    min = 0, 
    max = 1,
    value = 0.5)

)

renderPlot({
  
  create_PCA(data = raw_data, autoscale_input = input$autoscaler, transform_input = input$transformer_section_1, coloring_scheme_input = input$coloring_scheme_1, point_size_input = input$point_size_pca_1, opacity_input = input$opacity_pca_1, sample_label_input = input$sample_label_pca_1, corrected_data = FALSE)

})


```

---

### Metabolite Control Charts

The control charts below should be used to help identify inter- and intra- batch effects which may be correctable in the following section. Red hashed lines represent 95% confidence intervals while the black solid line represents the mean of the data.

```{r qc, echo=FALSE}
inputPanel(
  
  varSelectInput(
  inputId = "cc_metabolite",
  label = "Metabolite Selection:",
  data = raw_data[, 5:ncol(raw_data)],
  selected = NULL,
  multiple = FALSE,
  selectize = TRUE,
  width = NULL,
  size = NULL
),

selectInput(
  inputId = "cc_class_filter", 
  label = "Class Selection:", 
  choices = c("Both", "QC", "Sample"), 
  selected = "Sample"),

sliderInput(inputId = "cc_point_size", 
            label = "Point Size:",
            min = 1, 
            max = 10,
            value = 5)

)

# plot control chart

renderPlot({
  
  create_control_chart(data = raw_data, metabolite_input = input$cc_metabolite, class_filter_input = input$cc_class_filter, point_size_input = input$cc_point_size, transform_input = input$transformer_section_1, corrected_data = FALSE)
  
})


```

<hr style="border:1px solid black">

<div align = "center">

## Section 2: Batch Correction and Post-Correction Data Summaries

</div>

<div align = "left">

<hr style="border:1px solid black">

### Batch Correction Algorithm Selection

The following batch correction algorithms are supported for implementation:

1. [Quality Control - Robust Spline Correction (QC-RSC)](https://link.springer.com/article/10.1007/s00216-013-6856-7): This algorithm has a user definable smoothing parameter (0 < p < 1), however, to help avoid overfitting, p can be set to 0. In this case, it will be automatically optimized using leave-one-out cross validation. The QC-RSC algorithm is built upon the [quality control - robust LOESS signal correction (QC-RLSC)](https://www.nature.com/articles/nprot.2011.335) algorithm.

2. Quality Control - Median Normalization (QC-MN): This is a simple correction algorithm which simply normalizes each metabolite response to the median of the nearest (n) QC responses. For instance, if n is set to 3 then metabolite A in sample 1 will be normalized to the median response of metabolite A from QCs 1-3. Median is used to reduce the impact of outliers. A larger n should also reduce the impact of outliers but may include QCs less reflective of the local batch effects. More information on this correction algorithm is provided by [Martens et al.](https://www.mdpi.com/2218-1989/13/5/665). 

The selected data transformation is applied following batch correction to all of section 2.

```{r}

inputPanel(
  
  selectInput(
  inputId = "batch_correction", 
  label = "Batch Correction Algorithm:", 
  choices = c("QC-RSC", "QC-MN"), 
  selected = "QC-RSC"),
  
  sliderInput(inputId = "smoothing_parameter", 
            label = "Smoothing Parameter (QC-RSC):",
            min = 0, 
            max = 1,
            value = 0),
  
  sliderInput(inputId = "n_qc", 
           label = "Number of QCs (QC-MN):",
           min = 2, 
           max = 10,
           value = 3),
  
  selectInput(
  inputId = "transformer_section_2",
  label = "Data Transformer:",
  choices = c("None", "Log10", "ln", "Cube Root"),
  selected = "None",
  multiple = FALSE,
  selectize = TRUE,
  width = NULL,
  size = NULL)
  
)

```

### Impact of Correction on Technical and Biological Precison 

The following plot summarizes the improvement the batch-correction had on technical variation (select "QC" under "Class Selection") and biological variation (select "Sample" under "Class Selection"). The x-axis "CV-Fold Change" is a ratio of the uncorrected CV to the corrected CV. Thus a value greater than one indicates a reduction in variation. The y-axis "Post-Correction CV" indicates the variation following correction. The option to include labels can be used to help identify outliers following correction.

```{r}

inputPanel(
  
    selectInput(
    inputId = "volcano_class_filter", 
    label = "Class Selection:", 
    choices = c("Both", "QC", "Sample"), 
    selected = "QC"),

    sliderInput(
    inputId = "point_size_volcano", 
    label = "Point Size:",
    min = 1, 
    max = 10,
    value = 4),
  
    sliderInput(
    inputId = "opacity_volcano", 
    label = "Opacity:",
    min = 0, 
    max = 1,
    value = 0.5),

    selectInput(
    inputId = "sample_label_volcano",
    label = "Sample Labels:",
    choices = c("TRUE", "FALSE"),
    selected = "FALSE",
    multiple = FALSE,
    selectize = TRUE,
    width = NULL,
    size = NULL)

)

renderPlot({
  
create_volcano(data = raw_data, transform_input = input$transformer_section_2, class_filter_input = input$volcano_class_filter, point_size_input = input$point_size_volcano, opacity_input = input$opacity_volcano, sample_label_input = input$sample_label_volcano)[[1]]
  
})

renderText({
  
create_volcano(data = raw_data, transform_input = input$transformer_section_2, class_filter_input = input$volcano_class_filter, point_size_input = input$point_size_volcano, opacity_input = input$opacity_volcano, sample_label_input = input$sample_label_volcano)[[2]]
  
})

```


### Post-Correction Table Summary

```{r}
# initialize reactive elements

table_summary_values <- reactiveValues(
  bio_cv_median = NULL,
  bio_cv_range = NULL,
  tech_cv_median = NULL,
  tech_cv_range = NULL
)

# Input Table

renderDataTable({
  
create_table(data = raw_data, transform_input = input$transformer_section_2, corrected_data = TRUE)[[1]]
  
})

renderText({
  
  create_table(data = raw_data, transform_input = input$transformer_section_2, corrected_data = TRUE)[[2]]
  
})

```


### Post-Correction PCA

```{r}

inputPanel(
  
    selectInput(
    inputId = "autoscaler_2",
    label = "Autoscale Data:",
    choices = c("TRUE", "FALSE"),
    selected = "TRUE",
    multiple = FALSE,
    selectize = TRUE,
    width = NULL,
    size = NULL),
    
    selectInput(
    inputId = "coloring_scheme_2",
    label = "Color By:",
    choices = c("Class", "Batch"),
    selected = "Class",
    multiple = FALSE,
    selectize = TRUE,
    width = NULL,
    size = NULL),
  
    selectInput(
    inputId = "sample_label_pca_2",
    label = "Sample Labels:",
    choices = c("TRUE", "FALSE"),
    selected = "FALSE",
    multiple = FALSE,
    selectize = TRUE,
    width = NULL,
    size = NULL),
  
  sliderInput(
    inputId = "point_size_pca_2", 
    label = "Point Size:",
    min = 1, 
    max = 10,
    value = 3),
  
    sliderInput(
    inputId = "opacity_pca_2", 
    label = "Opacity:",
    min = 0, 
    max = 1,
    value = 0.5)

)

renderPlot({
  
  create_PCA(data = raw_data, autoscale_input = input$autoscaler_2, coloring_scheme_input = input$coloring_scheme_2, point_size_input = input$point_size_pca_2, opacity_input = input$opacity_pca_2, transform_input = input$transformer_section_2, sample_label_input = input$sample_label_pca_2, corrected_data = TRUE)

})


```

### Post-Correction Control Charts

```{r}
inputPanel(
  
  varSelectInput(
  inputId = "cc2_metabolite",
  label = "Metabolite Selection:",
  data = raw_data[, 5:ncol(raw_data)],
  selected = NULL,
  multiple = FALSE,
  selectize = TRUE,
  width = NULL,
  size = NULL
),

selectInput(
  inputId = "cc2_class_filter", 
  label = "Class Selection:", 
  choices = c("Both", "QC", "Sample"), 
  selected = "Sample"),

sliderInput(inputId = "cc2_point_size", 
            label = "Point Size:",
            min = 1, 
            max = 10,
            value = 5)

)

# plot control chart

renderPlot({
  
  create_control_chart(data = raw_data, metabolite_input = input$cc2_metabolite, class_filter_input = input$cc2_class_filter, point_size_input = input$cc2_point_size, transform_input = input$transformer_section_2, corrected_data = TRUE)
  
})

```

<hr style="border:1px solid black">

<div align = "center">

## Section 3: Data Transformations and Filtering

</div>

<div align = "left">

<hr style="border:1px solid black">

### Distribution Transformation

The selected data transformation is applied to all of section 3.

```{r}
inputPanel(
  
  selectInput(
  inputId = "transformer_section_3",
  label = "Data Transformer",
  choices = c("None", "Log10", "ln", "Cube Root"),
  selected = "None",
  multiple = FALSE,
  selectize = TRUE,
  width = NULL,
  size = NULL)
  
)

```

### Metabolite Distribution Overview

This section explores the distribution of the data following batch correction. 

```{r}
inputPanel(
  
  varSelectInput(
  inputId = "metabolite_h",
  label = "Metabolite Selection:",
  data = raw_data[, 5:ncol(raw_data)],
  selected = NULL,
  multiple = FALSE,
  selectize = TRUE,
  width = NULL,
  size = NULL),

selectInput(
  inputId = "class_filter_h", 
  label = "Class Selection:", 
  choices = c("Both", "QC", "Sample"), 
  selected = "Sample"),

numericInput(
  inputId = "bins_h",
  label = "Number of Bins (Histogram):",
  value = 15,
  min = 0,
  max = NA,
  step = 1,
  width = NULL),

)

# Histogram

output$histogramPlot <- renderPlot({
  
  create_histogram(data = raw_data, transform_input = input$transformer_section_3, metabolite_input = input$metabolite_h, bin_input = input$bins_h, class_filter_input = input$class_filter_h, corrected_data = TRUE)
  
})

# Quantile-quantile plot

output$qqPlot <- renderPlot({
  
  create_qq(data = raw_data, transform_input = input$transformer_section_3, metabolite_input = input$metabolite_h, class_filter_input = input$class_filter_h, corrected_data = TRUE)
  
})

# Metabolite Histograms and Q-Q Plots

renderUI({
  fluidRow(
    column(6, plotOutput("histogramPlot")),
    column(6, plotOutput("qqPlot"))
  )
})


```

### Data Filtering and Export

Filter the final feature table using the following criteria:

1. Technical Variation (%)
2. Intraclass correlation Coefficient 

Provide a file name for a csv of the final data set. The data exports automatically.

```{r}
inputPanel(
  
  numericInput(
  inputId = "qc_cv_cutoff",
  label = "Technical Precision Threshold:",
  value = 40,
  min = 0,
  max = NA,
  step = 0.01,
  width = NULL),
  
  numericInput(
  inputId = "icc_cutoff",
  label = "ICC Threshold:",
  value = 0.7,
  min = 0,
  max = 1,
  step = 0.01,
  width = NULL),
  
  textInput(
    inputId = "csv_name", 
    label = "CSV Name:", 
    value = "Corrected Data Matrix", 
    width = NULL, 
    placeholder = NULL)
  
)

renderText({
paste("Metabolites with a technical CV above ", input$qc_cv_cutoff, "%:", sep = "")
})

renderText({
filter_function(data = raw_data, transform_input = input$transformer_section_3, corrected_data = TRUE, qc_cv_threshold = input$qc_cv_cutoff, icc_threshold = input$icc_cutoff)[[1]]
})

renderText({
paste("Metabolites with an ICC below ", input$icc_cutoff, ":", sep = "")
})

renderText({
filter_function(data = raw_data, transform_input = input$transformer_section_3, corrected_data = TRUE, qc_cv_threshold = input$qc_cv_cutoff, icc_threshold = input$icc_cutoff)[[2]]
})

reactive({
  
export_function(data = raw_data, transform_input = input$transformer_section_3, corrected_data = TRUE, qc_cv_threshold = input$qc_cv_cutoff, icc_threshold = input$icc_cutoff, csv_name = input$csv_name)
  
})

```

